---
title: "Credit Risk Assessment"
author: "Ubayd Knight"
date: "`r Sys.Date()`"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: united
runtime: shiny
---

## Column

### Guide

**Welcome to the Home Loan Default Risk Predictor App!**

This app helps you assess the risk of loan default based on various user inputs. It uses a logistic regression model to predict the probability of default.

**How to Use the App:**

1.  Enter the loan amount **(must be between £1,100 and £110,000)**.

2.  Provide the mortgage due amount **(must be non-negative)**.

3.  Enter the property value **(must be non-negative)**.

4.  Select the reason for the loan **(Home Improvement or Debt Consolidation)**.

5.  Choose your job category **(Manager, Office, Other, Professional/Executive, Sales, Self-Hired)**.

6.  Enter the years at your present job **(must be between 0 and 100 years)**.

7.  Enter the number of major derogatory reports **(must be between 0 and 10)**.

8.  Enter the number of delinquent credit lines **(must be between 0 and 19)**.

9.  Enter the age of your oldest credit line in months **(must be between 0 and 100)**.

10. Enter the number of recent credit inquiries **(must be between 0 and 100)**.

11. Enter the number of credit lines **(must be between 0 and 100)**.

12. Enter your debt-to-income ratio **(must be between 0 and 100)**.

Click the **"Predict Default Risk"** button to see the results.

**Methodology:** The app uses a logistic regression model trained on historical loan data to predict the probability of default. The model considers various factors such as loan amount, mortgage due, property value, job category, years at present job, number of derogatory reports, number of delinquent credit lines, age of oldest credit line, number of recent credit inquiries, number of credit lines, and debt-to-income ratio.

**Model Details:** The logistic regression model is used to predict the probability of default. Key features include loan amount, mortgage due, property value, job category, years at present job, number of derogatory reports, number of delinquent credit lines, age of oldest credit line, number of recent credit inquiries, number of credit lines, and debt-to-income ratio.

**Interpretation of Results:** The bar output shows the predicted probability of default. The bar is divided into three colors:

-   **Green (0-35%)**: Low risk of default.

-   **Yellow (36-64%)**: Moderate risk of default.

-   **Red (65-100%)**: High risk of default.

**Acknowledgments:** This app was developed with help from the YaRrr! Pirate's Guide to R and the inbuilt R documentation.

In the "hmeq.csv" dataset, the results for the reasoning of the home equity loans as well as the job options have been renamed to their full version (e.g. ProfExe -\> Professor/Executive) for user convenience.

**Data Source:** The model is trained on a dataset named "hmeq.csv," which includes historical loan data with various features. The dataset is sourced from creditriskanalytics.net. As a token of appreciation I will include their books as references below.

**References:**

-   B. Baesens, D. Roesch, H. Scheule, *Credit Risk Analytics: Measurement Techniques, Applications and Examples in SAS*, Wiley, 2016.

-   H. Scheule, D. Roesch, B. Baesens, *Credit Risk Analytics: The R Companion*, Scheule Roesch Baesens, 2017.

**Contact Information:** For any questions, suggestions, or support, please contact me at u2281887\@unimail.hud.ac.uk.

## Column {data-width="700"}

### Application

```{r}
```

------------------------------------------------------------------------

```{r}
#
# This is a Shiny web application made by Ubayd Knight. 
# You can run the application by clicking
# the 'Run App' button above.

# To help, I used the YaRrr! Pirate's Guide to R as learning material
# To access the book, please visit this link: https://bookdown.org/ndphillips/YaRrr/

# Since I knew that logistic regressions were widely used for circumstances
# such as this, I knew making this wouldn't seem as hard as it would for others.

# For the logistic regression, I referred to 15.5 to convert the numbers into
# probabilities.

# In the "hmeq.csv" dataset, the results for the reasoning of the home equity loans as well as the job options have been renamed to their # full version (e.g. ProfExe -\> Professor/Executive) for user convenience.

# Data Source: The model is trained on a dataset named \"hmeq.csv,\" which includes historical loan data with various features. The       # dataset is sourced from creditriskanalytics.net. As a token of appreciation I have included a reference to in GitHub.

# I also used the inbuilt R documentation, mainly for visualisation purposes
# regarding shiny, shinythemes, shinywidgets, etc.
#

library(shiny)
library(shinythemes)
library(shinyWidgets)
library(ggplot2)
library(caret)
library(dplyr)
library(flexdashboard)
library(shinyFeedback)
library(rsconnect)

ui <- fluidPage(
    theme = shinytheme("yeti"), #this sets the appearance using shinytheme

    fluidRow(
        column(12, align = "center", offset = 0.5, #telling the code to basically center the title
               titlePanel("Home Loan Default Risk Predictor"))
    ),

    fluidRow(
        column(12, offset = 0.5, align = "center", #telling the code to basically center the title
               
               numericInput("loan_amnt", "Loan Amount:", value = "", min = 1100, max = 110000), #setting up user inputs for the app
               
               numericInput("mortdue", "Mortgage Due:", value = "", min = 0),
               
               numericInput("value", "Property Value:", value = "", min = 0),
               
               selectInput("reason", "Reason for Loan:", choices = c("Home Improvement", "Debt Consolidation")),
               
               selectInput("job", "Job:", choices = c("Manager", "Office", "Other", "Professional/Executive", "Sales", "Self-Hired")),
               
               numericInput("yoj", "Years at Present Job:", value = "", min = 0, max = 100),
               
               numericInput("derog", "Number of Major Derogatory Reports:", value = "", min = 0, max = 10),
               
               numericInput("delinq", "Number of Delinquenct Credit Lines:", value = "", min = 0, max = 19),
               
               numericInput("clage", "Age of Oldest Credit Line (In Months):", value = "", min = 0, max = 100),
               
               numericInput("ninq", "Number of Recent Credit Inquiries:", value = "", min = 0, max = 100),
               
               numericInput("clno", "Number of Credit Lines:", value = "", min = 0, max = 100),
               
               numericInput("debtinc", "Debt-to-Income Ratio:", value = "", min = 0, max = 100),
               
               actionButton("prediction", "Predict Default Risk")
  
        ),

        fluidRow(
            column(12, align = "center", #telling the code to basically center the prediction button
                   
                   plotOutput("predictionPlot") #telling the code to plot the "predictionPlot" which is made below
            )
        )
    )
)

server <- function(input, output) {
    loan_data <- read.csv("hmeq.csv") #telling the code read from the hmeq.csv file in the folder
  
    loan_data <- na.omit(loan_data) #telling the code to leave out entire rows that have missing information to keep the results
    # statistically valid
    
    loan_data$BAD <- as.factor(loan_data$BAD) #by inputting "as.factor" instead of "as.numeric", you are telling the code that this
    # represents categorical data instead of numerical data (0 for no default, and 1 for default, in this instance.)
  
    loan_data <- loan_data %>% #utilises the 'pipeline operator' from dplyr package basically saying it'll be used in other functions
      
        mutate( #this "mutates" or converts these columns into different data types, here, being numeric and factor
          
            LOAN = as.numeric(LOAN),
            
            MORTDUE = as.numeric(MORTDUE),
            
            VALUE = as.numeric(VALUE),
            
            REASON = as.factor(REASON),
            
            JOB = as.factor(JOB),
            
            YOJ = as.numeric(YOJ),
            
            DEROG = as.numeric(DEROG),
            
            DELINQ = as.numeric(DELINQ),
            
            CLAGE = as.numeric(CLAGE),
            
            NINQ = as.numeric(NINQ),
            
            CLNO = as.numeric(CLNO),
            
            DEBTINC = as.numeric(DEBTINC)
        )
  
    preProc <- preProcess(loan_data, method = c("center", "scale")) #applies consistency throughout the dataset by making the mean of
    # each function/variable zero so that the values are 'normalised' and a better fit for utilisation
    
    loan_data <- predict(preProc, loan_data) #applies the "preproc" instruction to the "loan_data" above
  
    train_control <- trainControl(method = "cv", number = 10) #this is a cross-validation technique to prevent overfitting and help make 
    # the model's performance become more reliable. The number 10 represents how many times the model will be trained
    
    model <- train(BAD ~ LOAN + MORTDUE + VALUE + REASON + JOB + YOJ + DEROG + DELINQ + CLAGE + NINQ + CLNO + DEBTINC, data = loan_data, method = "glm", family = "binomial", trControl = train_control) #assigns all the factors into the model, it basically is a formula for the
    # logistic regression model
  
    observeEvent(input$prediction, { #basically tells the code to make sure that it collects all the input values, preprocesses them,
      #makes a prediction, to be displayed as a result in the bar plot.
      
        new_data <- data.frame(
          
            LOAN = input$loan_amnt,
            
            MORTDUE = input$mortdue,
            
            VALUE = input$value,
            
            REASON = input$reason,
            
            JOB = input$job,
            
            YOJ = input$yoj,
            
            DEROG = input$derog,
            
            DELINQ = input$delinq,
            
            CLAGE = input$clage,
            
            NINQ = input$ninq,
            
            CLNO = input$clno,
            
            DEBTINC = input$debtinc
        )
    
        new_data <- predict(preProc, new_data)
    
        print(new_data)
    
        if (any(is.na(new_data))) { #error in case the user accidentally (or purposely lol) leaves out a requirement
          
            showNotification("Please fill in all fields.", type = "error")
          
            return(NULL)
        }
    
        if (nrow(new_data) != 1) { #this is a debug instruction that helped me save time and hair in the process of making this app
          
            showNotification("Error: Incorrect number of rows in new_data.", type = "error")
          
            return(NULL)
        }
    
        if (!is.data.frame(new_data)) { #this is a debug instruction that helped me save time and hair in the process of making this app
          
            showNotification("Error: new_data is not a data frame.", type = "error")
          
            return(NULL)
        }
    
        prediction <- tryCatch({
          
            predict(model, new_data, type = "prob")[,2]
          
        }, error = function(e) { # this is a debug instruction that helped me save time and hair in the process of making this app
          
            showNotification(paste("Prediction error:", e$message),
                             
                             type = "error")
          
            return(NULL)
        })
    
        print(prediction)
    
        if (!is.null(prediction)) {
          
            output$predictionPlot <- renderPlot({ #renders the plot named "predictionPlot" and outputs it into the user interface
              
                color <- ifelse(prediction <= 0.35, "green", ifelse(prediction <= 0.64, "yellow", "red")) #changes bar colour depending
                # on the level of risk
              
                ggplot(data.frame(Prediction = prediction), aes(x = "", y = Prediction)) +
                
                    geom_bar(stat = "identity", fill = color, width = 0.4) + #geometry stuff to ensure the plot works with geo settings
                
                    ylim(0, 1) +
                
                    geom_text(aes(label = round(Prediction * 100, 2)), vjust = -0.5, size = 5) + #adds text label above the bar
                
                    labs(title = "", y = "Probability of Default", x = "") + #adds label and title to the graph
                  # to make it look nice and tidy :)
                
                    theme_minimal() + #minimal theme
                
                    theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
                          
                          axis.text.x = element_blank(), #removes x-axis text
                          
                          axis.ticks.x = element_blank()) #removes x-axis ticks
              
            })
        }
    })
}

shinyApp(ui = ui, server = server)
```
